---
title: "R final project report"
author: "Urban Merhar"
date: "11.1.2022"
output: pdf_document
---

[comment]: # (End of title page.)
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r project_code, include=FALSE}
source("R_final_project.R")
```

# Stochastic random process

For the final project in *Asset Pricing Theory* course I decided to simulate stock prices using *geometric Brownian motion*.

GBM explanation + economic explanation...

```{r Zt}
# stock_prices <- vector of stock prices in the past
# returns Z.t for calculation of mu.hat and sigma.hat
GBMmodel <- function(stock_prices){
  Z.t <- c()
  for(price in stock_prices){
    for(j in c(1:length(price)-1)){
      Z.t <- append(Z.t, log(price[j]/price[j+1]))
    }
  }
  return(Z.t)
}
```

```{r hats}
# calculate Z.t, mu.hat, sigma.hat
# df[2] is the vector of stock prices from my data (look at chapter Data)
Z.t <- GBMmodel(df[2])
mu.hat <- 1/(nrow(df) - 1) * sum(Z.t)
sigma.hat <- sqrt(var(Z.t)*1)
```

```{r prediction_price}
# returns a vector of predictions for the future using GBM
# mu.hat
# sigma.hat
GBMprediction <- function(mu.hat, sigma.hat){
  prediction.year <- 1:365
  predicted.price <- c()
  for(i in prediction.year){
    if(length(predicted.price) < 1){
      last.price <- df$Price[1]
    } else {
      last.price <- predicted.price[length(predicted.price)]
      }
    predicted.price <- append(predicted.price,
                               last.price * exp((mu.hat - sigma.hat^2 / 2) * 1/365 + sigma.hat * rnorm(1) * sqrt(1/365)))
  }
  return(predicted.price)
}
```

```{r simulated_data}
# Create simulated data
# returns a data frame of N simulations
#######################################
# N <- number of simulations
# start <- date of start of simulation (default "2022-1-11")
# end <- end date of simulation (default "2023-01-10")
# mu.hat
# sigma.hat
# date year-month-day
GBM.simulated.data <- function(N, start = "2022-1-11", end = "2023-01-10", mu.hat, sigma.hat){
  sim.df <- data.frame(future.date = seq(as.Date("2022-1-11"), as.Date("2023-01-10"), by="days"))
  for(i in 1:N){
    new <- as.data.frame(GBMprediction(mu.hat, sigma.hat))
    sim.df[ , ncol(sim.df) + 1] <- new
    colnames(sim.df)[ncol(sim.df)] <- paste0("sim", i)
  }
  return(sim.df)
}
```


# Data

The data was aquired on *11.1.2022* on the website of [nasdaq](https://www.nasdaq.com/). For the project I choose the stock of [Pfizer, Inc. Common Stock](https://www.pfizer.com/) as the company is very relevant in this covid times.

Historical prices of the pfizer stock can be found on the website [nasdaq](https://www.nasdaq.com/market-activity/stocks/pfe/historical). I downloaded data for the Pfizer, Inc. Common Stock for the last *5* years in *.csv* format and used the package *readr* to read the downloaded data. Bellow are the first six rows of raw data displayed with the *head* function.

```{r raw_data}
head(raw.data)
```

We have 6 columns:

1. `Date` date of trading,
2. `Close/Last` last price,
3. `Volume` volume of stocks,
4. `Open` open price,
5. `High` high price,
6. `Low` low price.

For the simulation i only need the `Date` and `Close/Last` as I decided to use the latter for the price of the stock.

The data was imported using `read_csv` function. After that I narrowed the columns to `Date` and `Price`. The last thing in cleaning the data was to get rid of the `$` sign and change the class of `Price` from character to double.

```{r data_import_and_cleaning}
# Pfizer stock data for the last 5 years
raw.data <- read_csv("HistoricalData_1641895850651.csv",
                     col_types = cols(Date = col_datetime(format = "%m/%d/%Y"), 
                                  "Close/Last" = col_character(), Volume = col_integer(), 
                                  Open = col_character(), High = col_character(), 
                                  Low = col_character()))

# For the price of stock I will take "Close/Last"
df <- raw.data[1:2]
df <- rename(df, "Price" = "Close/Last")

# get rid of $ and change char to double
df$Price <- as.numeric(gsub("\\$", "", df$Price))
```

The end result of cleaning the data is displayed bellow.

```{r data}
head(df)
```

After acquiring and cleaning the data we can visualize it. For visualizing I used the *ggplot2* package.

```{r graph_of_data}
# visualize the time series of selected data
ggplot(data = df, aes(x = Date, y = Price)) + 
  geom_line(color="navyblue") +
  labs(title = "Price of Pfizer stock", 
       subtitle = "Time series for the last 5 years", 
       caption = "Source: Nasdaq", 
       y = "Price USD") +
  scale_x_datetime(date_labels = "%Y",
                   date_breaks = "1 year") +
  theme_bw()
```

# Simulation results

Now that we have the code and the data we can simulate the price of stock using *geometric Brownian motion*.

```{r simulation}
# calculate Z.t, mu.hat, sigma.hat
Z.t <- GBMmodel(df[2])
mu.hat <- 1/(nrow(df) - 1) * sum(Z.t)
sigma.hat <- sqrt(var(Z.t)*1)

# simulate once
predicted.price <- GBMprediction(mu.hat, sigma.hat)

# simulate data
simulated.df <- GBM.simulated.data(50, mu.hat = mu.hat, sigma.hat = sigma.hat)
```

Now that we have simulated some data we should plot it. Lets plot `predicted.price` first, so we can see one simulation cleanly.

```{r first}
# visualize predicted price
ggplot() +
  geom_line(aes(x = 1:365, y = predicted.price),
            color = "navyblue") +
  labs(title = "GBM simulated price of Pfizer stock", 
       subtitle = "Simulated once",
       y = "Price USD",
       x = "Days") +
  theme_bw()
```

To show that we will get a different simulation if we run the code again lets make a second `predicted.price` and plot it again. 

```{r second}
# run a second simulation
predicted.price <- GBMprediction(mu.hat, sigma.hat)

# visualize predicted price
ggplot() +
  geom_line(aes(x = 1:365, y = predicted.price),
            color = "navyblue") +
  labs(title = "GBM simulated price of Pfizer stock", 
       subtitle = "Simulated once (second run)",
       y = "Price USD",
       x = "Days") +
  theme_bw()
```

Now lets see the graph of `simulated.df` where we simulated with *N = 50*.

```{r N_at_once}
# graph of N simulations

# for loop didn't work for graphs... :(
# help for plotting multiple lines with multiple colors in one go
# help from: https://www.geeksforgeeks.org/how-to-create-a-plot-using-ggplot2-with-multiple-lines-in-r/ 
library("reshape2")
data_long <- melt(simulated.df, id = "future.date", variable.name = "Simulation")
ggplot(data_long,            
       aes(x = future.date,
           y = value,
           color = Simulation)) +
  geom_line() +
  labs(title = "Simulation of Pfizer stock", 
       subtitle = "Simulations using Geometric Brownian motion", 
       y = "Price USD",
       x = "Date",
       name = "Simulation") +
  theme_bw()
```

```{r all_together}
# graph of past data + simulations
ggplot() +
  geom_line(data = data_long, aes(x = as.Date(future.date), y = value, color = Simulation)) +
  geom_line(data = df, aes(x = as.Date(Date), y = Price), color="navyblue") +
  labs(title = "Price of Pfizer stock and simulations", 
       subtitle = "Simulations using Geometric Brownian motion", 
       caption = "Source: Nasdaq", 
       y = "Price USD",
       x = "Date",
       name = "Simulation") +
  theme_bw() +
  theme(legend.position = "none")
```

